---
title: "Singapore HDB Resale Flat Prices for 1990 - 2022"
output: 
  github_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
```

## Introduction

As a millenial growing up in Singapore and starting out in my first job, one of my life's goals is to own a house on this little red dot with my life partner (*read: wife*). 

Here in [Singapore](https://en.wikipedia.org/wiki/Singapore), there are generally [two ways](https://www.propertyguru.com.sg/property-guides/the-different-types-of-housing-in-singapore-9916) to do this:

- [Public Housing](https://www.hdb.gov.sg/about-us/our-role/public-housing-a-singapore-icon)
  - Brand new Built-to-Order (BTO) Flats
  - Resale Flats
- Private Housing (e.g., condominiums, landed property, etc.)

Whilst BTOs tend to be the choice of most young Singaporean couples, my current and potential future situation may push my partner and I to look towards resale flats to meet our housing needs. As such, I set forth on this project thatread_csv("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices/Data/resale-flat-prices-based-on-approval-date-1990-1999") will focus on resale flats under the Housing & Development Board (HDB).

This project therefore seeks to analyse the prices of Singapore's HDB Resale Flats for the years of 1990 - 2022. Data was obtained from Kaggle, [(1)](https://www.kaggle.com/datasets/syrahmadi/resale-hdb-flat-prices-2000-2022) and [(2)](https://www.kaggle.com/datasets/chngyuanlongrandy/hdb-prices-with-closest-mrt-distance), which were sourced from GovTech's open data portal, [data.gov.sg](https://data.gov.sg/dataset/resale-flat-prices).

#### Importing datasets

```{r import}
#change wd to dataset folder
setwd("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices/Data")

read_csv("resale-flat-prices-based-on-approval-date-1990-1999.csv")
```

