---
title: "Singapore HDB Resale Flat Prices for 1990 - 2022"
output: 
  github_document:
    toc: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
setwd("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices")
```

## Introduction

As a millenial growing up in Singapore and starting out in my first job, one of my life's goals is to own a house on this little red dot with my life partner (*read: wife*). 

Here in [Singapore](https://en.wikipedia.org/wiki/Singapore), there are generally [two ways](https://www.propertyguru.com.sg/property-guides/the-different-types-of-housing-in-singapore-9916) to do this:

- [Public Housing](https://www.hdb.gov.sg/about-us/our-role/public-housing-a-singapore-icon)
  - Brand new Built-to-Order (BTO) Flats (or apartments)
  - Resale Flats
- Private Housing (e.g., condominiums, landed property, etc.)

Whilst BTOs tend to be the choice of most young Singaporean couples, my current and potential future situation may push my partner and I to look towards resale flats to meet our housing needs. As such, I set forth on this project that will focus on resale flats under the Housing & Development Board (HDB).

This project therefore seeks to analyse the prices of Singapore's HDB Resale Flats for the years of 1990 - 2022. Data was obtained from Kaggle, [(1)](https://www.kaggle.com/datasets/syrahmadi/resale-hdb-flat-prices-2000-2022) and [(2)](https://www.kaggle.com/datasets/chngyuanlongrandy/hdb-prices-with-closest-mrt-distance), which were sourced from GovTech's open data portal, [data.gov.sg](https://data.gov.sg/dataset/resale-flat-prices).

#### Importing datasets

```{r import, message=FALSE}
#import data as objects
nineties_data <- read_csv("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices/Data/resale-flat-prices-based-on-approval-date-1990-1999.csv")
feb2012_data <- read_csv("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices/Data/resale-flat-prices-based-on-approval-date-2000-feb-2012.csv")
dec2014_data <- read_csv("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices/Data/resale-flat-prices-based-on-registration-date-from-mar-2012-to-dec-2014.csv")
dec2016_data <- read_csv("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices/Data/resale-flat-prices-based-on-registration-date-from-jan-2015-to-dec-2016.csv")
latest_data <- read_csv("C:/Users/iylia.hutta/OneDrive - Meltwater/Desktop/Data Stuff/Singapore HDB Flat Resale Prices/Data/resale-flat-prices-based-on-registration-date-from-jan-2017-onwards.csv")
```


## Data Exploration

First, we will examine the data to understand its structure and contents.

```{r data exploration}
head(nineties_data)
head(feb2012_data)
head(dec2014_data)
head(dec2016_data)
head(latest_data)
```

From these, we see that the data structure is generally similar. The only difference being that the "Jan 2015 - Dec 2016" and "Jan 2017 onwards" data consists of an extra column `remaining_lease`, indicating the remaining years of lease for the houses in the datasets. For the purpose of this project then, we will be dropping this column to maintain data uniformity.

```{r drop excess columns}
dec2016_data <- dec2016_data %>% 
  select(-remaining_lease)

head(dec2016_data)

latest_data <- latest_data %>% 
  select(-remaining_lease)

head(latest_data)
```

Now that the data is similar, we can combine the datasets together:

```{r union datasets}
resale_data <- bind_rows(nineties_data,feb2012_data,dec2014_data,dec2016_data,latest_data)

head(resale_data)
```

Looking into the dataset, we see that the `storey_range` variable has varied values, which we will be renaming into a new variable called `storey`:

```{r examine storey_range}
resale_data %>% 
  group_by(storey_range) %>% 
  summarise(n())
```

However, before we create the new variable, it would be helpful to understand the distribution of storey ranges among the resale flats. This is because **most** old HDB flats (circa prior to 2010) were only built up to about 15 storeys.

```{r storey_ranges by year, warning = FALSE, message = FALSE}
library(do)

#Create year variable
resale_data <- resale_data %>% 
  mutate(year = as.integer(left(month, 4))) %>% 
  relocate(year, .before = town)

#Plot box plots of counts grouped by storey_ranges and faceted by year
#1990s HDB flats
resale_data %>% 
  filter(year < 2000) %>% 
  group_by(year, storey_range) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(x = storey_range, y = count)) +
  geom_boxplot() +
  coord_flip() +
  facet_wrap(~year)

#2000s HDB flats
resale_data %>% 
  filter(year >= 2000 & year < 2010) %>% 
  group_by(year, storey_range) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(x = storey_range, y = count)) +
  geom_boxplot() +
  coord_flip() +
  facet_wrap(~year)

#2010 - 2022 HDB flats
resale_data %>% 
  filter(year >= 2010 & year <= 2022) %>% 
  group_by(year, storey_range) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(x = storey_range, y = count)) +
  geom_boxplot() +
  coord_flip() +
  facet_wrap(~year)
```


## Exploratory Data Analysis





